\documentclass{article}
\usepackage[margin=2cm]{geometry}
\usepackage[dvips]{graphicx}
\begin{document}

\section*{Learning goals}
\label{sec:learning-goals}

Before the next day, you should have achieved the following learning
goals: 

\begin{itemize}
\item Use the main classes from package \verb+java.util.concurrent+.
\item Understand the concepts of \emph{thread pool} and \emph{graceful
  degradation}.
\end{itemize}

\section{Graceful degradation everywhere}
\label{sec:grac-degr-everywh}

Think of three examples of concurrent applications that you use every
week and that benefit from graceful degradation, i.e. becoming
gradually more slow under high load rather than crashing unexpectedly.

Once you have thought of three examples, check them with one of your
colleagues. Have you thought of the same examples?

\section{Text loops re-executed}
\label{sec:text-loops}

Modify the code of the exercise ``Text Loops'' from last day to use
one of the \verb+Executor+ instead of plain threads. 


\section{Responsive UI (that degrades gracefully)}
\label{sec:responsive-ui-that}

\subsection{Implement Executor}
\label{sec:implement-executor}

Create your own implementation of \verb+Executor+. Look at the
documentation of the API of \verb+Executor+ for ideas. Your
implementation must be able to execute one \verb+Runnable+ at a time
(but you can make it run several tasks in paralel, see also
Exercise~\ref{sec:impl-exec-}) and must have a queue of pending
tasks. 

Try your implementation by exchaning the class that you used in your
solution for the former exercise by your new implementation of
\verb+Executor+. 

\subsection{Extend Executor}
\label{sec:extending-executor}

Create a class \verb+TimedTask+ that implements
\verb+Runnable+ where the \verb+run()+ method is only a call to
\verb+Thread.sleep()+ for a number of milliseconds set at creation
time. This number must be at most 1000 (if a higher number is given at
construction time, it must be capped at 1000). 

Now extend your implementation of \verb+Executor+ so that it has a
public method \verb+getMaxPendingTime()+ that returns the maximum
theoretical time needed to run all tasks in the queue (i.e~assuming
all of them last 1000ms).


\subsection{Use Executor}
\label{sec:use-new-executor}

Modify your solution of exercise ``Responsive UI'' so that: 

\begin{itemize}
\item It has two classes: one represents the application and one
  represents its users.
\item The former uses your implementation of \verb+Executor+ with a
  thread of pools instead of using plain threads.
\item The latter will use an \verb+Executor+ from Java 
  to have a lot of
  threads representing users, and the threads will programatically
  create new tasks instead of the human user doing it by hand,
  i.e. there is no need to ask the user to enter data at any point in
  this version of the exercise. Threads should create tasks as
  fast as the time they need to be run (e.g.~if a task that will make
  the thread sleep for one second, the next task should be created
  a second later). 
\item The ``application'' class should implement a method
  \verb+getMaxWait()+ that returns the (theoretical) time needed to
  execute all tasks already in the queue.
\item The ``users'' (threads) will ask for the waiting time
  before they assign a new task. If the wait is above a certain
  threshold (set at construction time for each ``user''), the ``user'' 
  should print ``The site is down! I will come back later...'' and
  then wait for a long time before sending 
  more tasks. 
\end{itemize}

Try your implementation with different numbers of users and see how
many users it can handle. 

% TODO: think this through next year. It is 02h00 now...
%
% \subsection{TimedRunnable (*)}
% \label{sec:extending-executor}

% Create an interface \verb+TimedRunnable+ that extends \verb+Runnable+
% with a method \verb+timeNeeded()+, that returns the time needed to run
% the tasks in milliseconds. Create also a class that implements
% \verb+TimedRunnable+ where the \verb+run()+ method is only a call to
% \verb+Thread.sleep()+ for a number of milliseconds set at creation
% time.

% Now extend your implementation of \verb+Executor+ so that it is able
% to execute both normal \verb+Runnable+ tasks and \verb+TimedRunnable+
% task. Both types of tasks should be stored in different queues. It
% must also implement a public method \verb+getExactTime()+ 
% that returns the sum of the time needed to run all
% \verb+TimedRunnable+ in the queue. 


\section{Self-ordering list (*)}
\label{sec:self-ordering-list--}

Re-implement your solutions for the exercise ``Self-ordering list'' by
using an \verb+Executor+. 

\section{Dining philosophers (*)}
\label{sec:dining-philosophers}

Re-implement your solutions for the exercise ``Self-ordering list'' by
using an \verb+Executor+. 

\section{Implementing Executor Services (**)}
\label{sec:impl-exec-}

Create your own implementation of \verb+Executor+,
\verb+ExecutorService+ and 
\verb+ScheduledExecutorService+, with a thread pool of $n$, 
where $n$ is set at construction time. Use your implementation in the
former exercises. 


\end{document}