
\documentclass{article}
\usepackage[margin=2cm]{geometry}
\begin{document}

\section*{Learning goals}
\label{sec:learning-goals}

Before the next day, you should have achieved the following learning
goals: 

\begin{itemize}
\item Create full-Java programs, without any reliance on Groovy. These
  programs must be completely object-oriented and launch from the main
  method of one of the classes.
\item Understand what the keyword \verb+static+ means and how it
  works. 
\item Understand how the \verb+do...while+ loop structure works and
  when to use it instead of a \verb+while+ loop.
\item Be able to create linked lists and travel through their
  elements. 
\end{itemize}

You should be able to finish most of non-star exercises in the lab. 
Remember that star exercises are more difficult. 
\textbf{Do not try them unless the other ones are clear to you}. 

\section{Instance counter}
\label{sec:instance-counter}

{\bf Note:} This exercise is designed to help you practice the meaning of
\verb+static+. However, remember that adding static fields to your
classes is usually a bad idea. Do not do it unless you have a very
good reason. Hint: never do it in Programming in Java. 

\vspace{1em}

Complete the example given in the notes with a class called
\verb+Spy+. Your class must have: 

\begin{itemize}
\item one and only one \verb+static+ variable, and \verb+int+ called
  \verb+spyCount+.
\item an instance variable of type \verb+int+ for the spy's ID.
\item a constructor method that receives
  the ID of the spy as an argument, increases
  \verb+spyCount(int)+ by one, 
  and prints on the screen the ID of this spy plus the total number of
  spies so far.
\item a \verb+die()+ method that prints on the screen ``Spy XX has
  been detected and eliminated'' (where XX is the spy's ID), 
  decrements the spy counter and prints
  on the screen the total number of spies so far. 
\item a main method in which several objects of class \verb+Spy+ are
  created and some of them killed (their method \verb+die()+ is
  called). 
\end{itemize}

Observe how the static variable is accessed by different objects both
to increment and to decrement it. 

\section{do \{practice\} while (!understood);}
\label{sec:do-practice-while}

Make a class that implements a method 
that reads a list of marks between 0 and 100 from the
user, one per line, and stops when the user introduces a -1. The
program should output at the end (and only at the end) how many marks
there were in total, how many were distinctions (70--100), how many
were passes (50--69), how many failed (0--49), and how many were
invalid (e.g. 150 or -3). \textbf{Use \texttt{readLine()} exactly once}. The
output may look similar to this example: 

\begin{verbatim}
    Input a mark: 13
    Input a mark: 45
    Input a mark: 63
    Input a mark: 73
    Input a mark: 101
    Input a mark: 45
    Input a mark: 18
    Input a mark: 92
    Input a mark: -1
    There are 7 students: 2 distinctions, 1 pass, 4 fails (plus 1 invalid). 
\end{verbatim}

\section{Singly-linked lists}
\label{sec:doubly-linked-lists}

Create a linked list following  the ``hospital and patients'' example
of the notes:

\begin{enumerate}
\item Create the linked list and add several elements to it
  (around 10 is fine). 
\item Go through the list printing out the content of
  each element.
\item Delete a couple of elements from the list. 
\item Print the elements in the list again. Check that the deleted
  elements have been deleted. Check that you can delete the first
  element too. 
\end{enumerate}

Hint: Note that you cannot delete the first element from inside the
list because you need to update the pointer
\verb+patientListStart+. To add or delete the first element of a list,
you must do it from outside the list. 

\section{Queue length}
\label{sec:queue-length}

Reuse the code of one of the dynamic list in the former exercise, and
add a method that calculates the length of the list (maybe you have
done this already). 

Is it easy to do this both iteratively and recursively? Try both ways
and see which is more natural for you. If one takes too long, try the
other way. 

\section{Doubly-linked lists}
\label{sec:doublyr-linked-lists}

A doubly-linked list is a dynamic list in which each element is
connected to two other elements instead of just one. The one before
and the one after it. In this exercise, you must create a
doubly-linked list using the ``hospital and patients'' example as a
starting point.

\begin{enumerate}
\item Create the doubly-linked list and add several elements to it
  (around 10 is fine). 
\item Traverse it forwards and backwards printing out the content of
  each element.
\item Delete a couple of elements from the list. 
\item Traverse it forwards and backwards printing out the content of
  each element.
\item Add a new element to the list. Try to delete an element that is
  NOT in the list. 
\item Traverse it forwards and backwards printing out the content of
  each element.
\end{enumerate}

Hint: First you need to add a new field to Patient for the pointer
going ``backwards''. Then you need to modify the \emph{add} and
\emph{delete} methods to make sure you do not have loose pointers. 

\section{Circular lists}
\label{sec:circular-lists}

A circular list is a dynamic list in which there is a beginning but
there is no end: the last element is pointing back to the first
element. Circular lists can be singly- or doubly-linked. In this
exercise, you must create a singly-linked circular list using the
``hospital and patients'' example as a starting point.

\begin{enumerate}
\item Create the circular list and add several elements to it
  (around 10 is fine). 
\item Traverse it forwards for one complete loop. While you go around
  the list, print out the content of each element. How do you know
  that you have reached the end of the list when you never find a
  \verb+null+ pointer? 
\item Delete a couple of elements from the list. 
\item Traverse it again printing out the content of
  each element.
\item Add a new element to the list. Try to delete an element that is
  NOT in the list. 
\item Traverse it again printing out the content of
  each element.
\end{enumerate}

\section{Queues (*)}
\label{sec:queues}

A queue is a dynamic structure that implements these methods: 

\begin{description}
\item[insert(...)] inserts an element at the beginning of the queue.
\item[retrieve(...) ] remove an element from the \emph{end} of the queue.
\item[size() ] returns the current size of the queue. 
\end{description}

Queues are heavily used in computing: communication
buffers, incoming request to web servers, read/write requests to hard
disks, etc. 

Implement a queue of integers. This could represent requests
to a hard drive to read from different sectors. Implement the three
methods listed above. 

Then write another class that creates a queue, and makes several
\verb+insert()+, \verb+retrieve()+, and \verb+size()+ calls. Check
that the values you get are consistent. The output of the program
could look like this: 

\begin{verbatim}
    There are 0 requests in the queue.
    Inserting request 5...
    Inserting request 8...
    Inserting request 12...
    There are 3 requests in the queue.
    Retrieving request 5... done. 
    Inserting request 13...
    There are 3 requests in the queue.
    Retrieving request 8... done. 
    Retrieving request 12... done. 
    There are 2 requests in the queue.
    Retrieving request 13... done. 
    etc...
\end{verbatim}

\section{Stacks (*)}
\label{sec:stack}

A stack is a dynamic structure that implements these methods: 

\begin{description}
\item[push(...)] inserts an element at the beginning of the stack.
\item[pop(...) ] remove an element from the \emph{beginning} of the stack.
\item[empty(...) ] returns true if there are no elements on the stack,
  false otherwise.
\end{description}

Stacks are heavily used in computing. The method call stack that
stores the variables for each method in a program is just one
well-known example. 

Implement a stack of integers. This could represent requests
to a hard drive to read from different sectors. Implement the three
methods listed above. 

Then write another class that creates a stack, and makes several
\verb+push()+, and \verb+pop()+ calls. Make sure you check the size of
the stack before popping elements out. Check
that the values you get are consistent. The output of the program
could look like this: 

\begin{verbatim}
    Pushing 5...
    Pushing 8...
    Pushing 12...
    Popping... it's a 12
    Pushing 13...
    Popping... it's a 13
    Popping... it's a 8
    Popping... it's a 5
    Stack is empty
\end{verbatim}

\section{From array to list}
\label{sec:from-array-string}

Create a static method that takes an array of integers and returns a
linked list of integers in the same order as the original array. You
can put this method in a class called \verb+ListUtilities+ so that you
can easily use it for the following exercises. 

Initialising an array with 15-20 elements is very easy with
curly-brackets-notation, so it is 
common to create an array and then convert it into a dynamic structure
automatically. 

\section{A sorted list}
\label{sec:sorted-list}

Create a linked list of integers that is automatically sorted. 

Everytime you add an element, make sure you introduce it at the right
place in the queue so that the queue is always sorted from lower to
higher numbers. 

Create the list, insert several numbers into it (around 15 is fine)
and then go through the list printing the value of each element. Check
that the elements are correctly sorted.


\section{Bubble sort (*)}
\label{sec:bubble-sort}

In this exercise, you will implement the \emph{bubble sort} algorithm
and use it to sort an unsorted linked list of integers. 

The bubble sort algorithm is the simplest sorting algorithm there is: 

\begin{enumerate}
\item Take the first element. Compare it with the second element. If
  it is greater, make them change positions (first becomes second and
  viceversa); otherwise, do nothing.
\item Move to the next element. Repeat the process (i.e.~compare 2 and
  3, then 3 and 4, up to the last element).
\item Now you have the highest element at the end of the list.
\item Move back to the beginning and repeat the whole process until
  your list is sorted.
\item Once you pass through the list without swapping elements, your
  list is sorted. 
\end{enumerate}

Put your implementation in a static method in a class, maybe
\verb+ListUtilities.bubbleSort(List)+. The name of the algorithm comes
from metaphorical bubbles always moving up in a liquid. 

Try your method with different lists and see how long it takes to sort
them. You can use the static method \verb+System.currentTimeMillis()+
to print on the screen the current time (measured in milliseconds
since 1st January 1970). 

\section{Cocktail sort (*)}
\label{sec:cocktail-sort}

Cocktail sort is a combination of two bubble sorts. First the list is
traversed in one direction to move the highest element to the end, and
then it is traversed in the opposite direction to move the lowest
element to the beginning. This bidirectional process is repeated until
there are no swaps in the list, which shows that the list is sorted. 

Implement cocktail sort in a static method in a class, maybe
\verb+ListUtilities.cocktailSort(List)+. The name of the algorithm comes
from a metaphorical cocktail shaker moving up and down, up and
down\ldots 

Try your method with the same lists as bubble sort 
and see how long it takes to sort them. 

\section{QuickSort (**)}
\label{sec:quicksort-}

This is a popular sorting algorithm that is very fast in most
situations. Once you have finished all the other exercises, look up in
wikipedia how this method works and make your own implementation (put
it in \verb+ListUtilities.quickSort(List))+. Then compare its
performance with the other two methods. In order to see significant
differences, you will need to use lists of some length. 


\end{document}